---
title: "Figure 4"
author: "Charlotte Uden"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(gridExtra)
```

Fig. 5. Observed versus predicted lightning strike rates. The dotted 1:1 line indicates perfect model performance. All axes are in units of lightning strikes/km2/month. The x-axis in each plot shows the observed strike rates, while the y-axes are predictions from (a) model N13, (b) model G13, and (c) model B13. 

Load configuration files
- location of the observed data (model drivers)
- location of table of model predictions exported from each model set's .Rmd file. 
- location to store the figure
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

config <- read_properties("/path/to/config/file/my_config.properties")  # path to your config file

# Extract values from config
data_root <- config[["data_root"]] 
table_root <- config[["table_root"]] 
figure_root <- config[["figure_root"]]

# Use the values to build full paths
Chen2021_file_path <- file.path(table_root, "Chen2021_models_predictions.csv")
lm_file_path <- file.path(table_root, "Linear_Model_predictions.csv")
glm_file_path <- file.path(table_root, "Gamma_GLM_predictions.csv")
bayes_file_path <- file.path(table_root, "bayes_gamma_model_predictions.csv")
monthly_drivers_file_path <- file.path(data_root, "monthly_drivers.csv")
```

```{r}
#read in the data
chen_predictions <- read.csv(Chen2021_file_path)[,2:6]
lm_predictions <- read.csv(lm_file_path)[,2:14]
glm_predictions <- read.csv(glm_file_path)[,2:14]
bayes_predictions <- read.csv(bayes_file_path)[,2:14]
```

Observed data - need to be split back into test and train sets because the above predictions are made from the test set. 
```{r}
data <- read.csv(monthly_drivers_file_path)
data <- data[2:13]
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```

Combine into one data frame. Note that N13, G13, and B13 are used because they combine all size near-surface variables and are the best performing models within each of their respective model sets. 
```{r}
df <- data.frame(
  Observed = test$strikes,
  N13 = lm_predictions$N13,
  G13 = glm_predictions$G13,
  B13 = bayes_predictions$y_pred_B13
)

head(df)
```

```{r}
p1 <- ggplot(data=df, aes(x=Observed, y=N13, alpha=0.7)) +
  geom_point() +
  xlim(0, 1.35) +
  ylim(-0.07, 1.35) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "black") +
  labs(title = "(a)",
       x = "Observed",
       y = "Predicted: N13") +
  guides(alpha = FALSE) +
  theme_minimal() +
  theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)) 
  
p1

```

```{r}
p2 <- ggplot(data=df, aes(x=Observed, y=G13, alpha=0.7)) +
  geom_point() +
  xlim(0, 1.35) +
  ylim(-0.07, 1.35) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "black") +
  labs(title = "(b)",
       x = "Observed",
       y = "Predicted: G13") +
  guides(alpha = FALSE) +
  theme_minimal() +
  theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)) 
  
p2

```

```{r}
p3 <- ggplot(data=df, aes(x=Observed, y=B13, alpha=0.7)) +
  geom_point() +
  xlim(0, 1.35) +
  ylim(-0.07, 1.35) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "black") +
  labs(title = "(c)",
       x = "Observed",
       y = "Predicted: B13") +
  guides(alpha = FALSE) +
  theme_minimal() +
  theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)) 
  
p3

```

```{r}
p1 | p2 |p3
ggsave(file.path(figure_root, "Figure_5.jpg"), plot = last_plot(), width = 12, height = 4, dpi = 750)
```
