---
title: "Figure 1"
author: "Charlotte Uden"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(ggpointdensity)
library(viridis)
library(terra)
```

Fig 1. Spatial and temporal distribution of lightning and climate data. Plots include (a) cloud-to-ground lightning strike rates (strikes/km2/month) across the Northeastern US for 2005 - 2010 plotted against month, and mapped (b) cloud-to-ground lightning strike rate (strikes/km2/month), (c) CAPE (J/kg), (d) CAPE x Precipitation (W/m2), (e) Temperature (Celsius), (f) Wind speed (m/s), (g) Short-wave radiation (W/m2), (h) Surface pressure (Pa), (i) Relative humidity (%), and (j) Precipitation (kg/m2/s). Mapped values are for summer months (May - September), averaged across the 2005-2010 training period. All variables excluding the target (lightning strike rate) are standardized. Lightning data are derived from the Vaisala Lightning Database and climate data come from the ERA5 climate reanalysis product (Vaisala, Inc.; Hersbach et al. 2020). 

Load configuration files
- location of climate and lighting drivers
- location to store the figure
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

config <- read_properties("/raid/cuden/config_files/lightning_config.properties")  # path to config file

# Extract values from config
data_root <- config[["data_root"]] #path to drivers
figure_root <- config[["figure_root"]] #path to wherever you want to save figures

# Use the values to build full paths
daily_drivers_file_path <- file.path(data_root, "era5DailySummaries_vaisalaLightningDailyCounts_2005-2010_NEclip.csv")
monthly_drivers_file_path <- file.path(data_root, "era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")
```


Read in daily strike count data:
```{r}
#daily data with lat lon and date
df <- read.csv(daily_drivers_file_path)
df <- df[,2:16]
```

In order to calculate strikes/km2/month from a daily strike count, you need to find the area of each gridcell:
```{r}
#get map of the Northeast
states <- map_data("state")#turn state line map into data frame
new_england <- subset(states, region %in% c("vermont", "new hampshire", "connecticut", "maine", "rhode island", "massachusetts", "new york"))#subest ne$
map <- ggplot() + geom_polygon(data = new_england, aes(x=long, y=lat, group = group), color="white", fill="lightgray") + coord_fixed(1.3)

#convert points to raster and set crs

#get grid
points <- filter(df, date=="2008-06-01")
map + geom_point(data=points, aes(x=lon, y=lat, size=cape, col=strikes))

#add unique id to each point
points$id <- seq(1:nrow(points))
#concert to raster 
r <- rast(points[,c("lon", "lat", "cape")])
plot(r)
crs(r) <- "+init=epsg:4326"
#now make a raster that includes the id column
r <- rast(points[,c("lon", "lat", "id")])
plot(r)
crs(r) <- "+init=epsg:4326"
#convert to polygons
p = as.polygons(r, aggregate=FALSE) #if aggregate=false, each raster cell is a polygon 
plot(p)    
#get area of each polygon
area <- expanse(p, unit="km", transform=T) #	If transform=TRUE, planar CRS are transformed to lon/lat for accuracy. can also use "m" instead of "km"
p$area <- area

plot(p, "area")
plot(p, "id")

d <- data.frame(values(p))
head(d)

#merge grid and area values by id
d <- merge(points[,c("lat", "lon", "id")], d, by="id")

#merge area values with data by lat lon
df <- left_join(df, d, by=c("lat", "lon"))

#map to check it worked
points <- filter(df, date=="2008-06-01")
map + geom_point(data=points, aes(x=lon, y=lat, size=cape, col=strikes))
map + geom_point(data=points, aes(x=lon, y=lat, col=as.factor(area)))

```
Calculate strike rate for each month:
```{r}
df$date <- as.Date(df$date)
df$month <- as.numeric(format(df$date,'%m'))
df$year <- as.numeric(format(df$date,'%Y'))
df$day <- as.numeric(format(df$date,'%d'))
df$week <- week(df$date)

cols <- c("lon", "lat", "year", "month", "area")

#sum strike count for each summer month 
summary_strikes <- df %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(strikes_monthly_sum = sum(strikes), .groups = 'drop')

#calculate strike rate for each month (km-2 month-1)
summary_strikes <- mutate(summary_strikes, strike_rate = strikes_monthly_sum/area)

print(summary_strikes)
```
Plot strike rates for each month:
```{r}
summary_strikes$strike_rate <- as.numeric(summary_strikes$strike_rate)

# Convert month number to factor with month abbreviations
summary_strikes <- summary_strikes %>%
  mutate(month_name = factor(month, levels = 1:12, labels = month.abb))

p1 <- ggplot(summary_strikes, aes(x = month_name, y = strike_rate)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1, color = "black") +
  geom_boxplot(outlier.shape = NA, width = 0.4, fill = "white", color = "white", alpha = 0.1) +  # no outliers shown
  labs(x = "", y = "Strikes / km² / month", title="(a)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    panel.grid.major = element_line(color = "black", size = 0.05), 
    panel.grid.minor = element_line(color = "black", size = 0.01),
    legend.position = "none"
  )

p1
```


Maps of each climate variable. First get the base map:
```{r}
#get map of the NE
library(mapdata)
# get new england and new york polygons from mapdata library :
states <- map_data("state")#turn state line map into data frame
northeast <- subset(states, region %in% c("vermont", "new hampshire", "connecticut", "maine", "rhode island", "massachusetts", "new york"))#subest northeastern states

# plot state polygons and PalEON grid, just to see what we are working with:
map <- geom_polygon(data = northeast, aes(x=long, y = lat, group = group), fill = NA, color = "black", size=0.2) 
```

Need the monthly summary data used to drive the models:
```{r}
#monthly averages
data <- read.csv(monthly_drivers_file_path)

#standardize the data
data <- data %>%
  mutate(cape = ((cape-mean(cape))/sd(cape))) %>%
  mutate(precip = ((precip-mean(precip))/sd(precip))) %>%
  mutate(cxp = ((cxp-mean(cxp))/sd(cxp))) %>%
  mutate(tair = ((tair-mean(tair))/sd(tair))) %>%
  mutate(wind = ((wind-mean(wind))/sd(wind))) %>%
  mutate(swr = ((swr-mean(swr))/sd(swr))) %>%
  mutate(sp = ((sp-mean(sp))/sd(sp))) %>%
  mutate(rh = ((rh-mean(rh))/sd(rh)))

#calculate mean values across years
cols <- c("lon", "lat")

#Lightning strike Rate
summary_strike <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(strikes = mean(strikes), .groups = 'drop')
```

Add each climate variable to the map:
```{r}
library(terra)
r <- rast(summary_strike[,c("lon", "lat", "strikes")])
plot(r)
crs(r) <- "+init=epsg:4326"

p_lightning <- ggplot() + 
  geom_raster(data = summary_strike, aes(x = lon, y = lat, fill = strikes)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Lightning  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) + ggtitle("(b)") + map +
  xlab("Lon") + ylab("Lat")

p_lightning


#CAPE
summary_cape <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(cape = mean(cape), .groups = 'drop')

r <- rast(summary_cape[,c("lon", "lat", "cape")])
plot(r)
crs(r) <- "+init=epsg:4326"

p_cape <- ggplot() + 
  geom_raster(data = summary_cape, aes(x = lon, y = lat, fill = cape)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "CAPE  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) + ggtitle("(c)") + map+
  xlab("Lon") + ylab("Lat")

p_cape

#CAPE x Precipitation
summary_cxp <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(cxp = mean(cxp), .groups = 'drop')

r <- rast(summary_cxp[,c("lon", "lat", "cxp")])
plot(r)
crs(r) <- "+init=epsg:4326"

p_cxp <- ggplot() + 
  geom_raster(data = summary_cxp, aes(x = lon, y = lat, fill = cxp)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "CAPE x Precip  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal") +
  ggtitle("(d)") + map+
  xlab("Lon") + ylab("Lat")
p_cxp

#Temperature
summary_temp <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(tair = mean(tair), .groups = 'drop')

p_temp <- ggplot() + 
  geom_raster(data=summary_temp, aes(x=lon, y=lat, fill=tair)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Temp  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal") +
  ggtitle("(e)") + map +
  xlab("Lon") + ylab("Lat")

p_temp


#Wind
summary_wind <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(wind = mean(wind), .groups = 'drop')

p_wind <- ggplot() + 
  geom_raster(data=summary_wind, aes(x=lon, y=lat, fill=wind)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Wind  ")) + 
  theme_minimal() +
  theme(
       axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) +
  ggtitle("(f)") + map+
  xlab("Lon") + ylab("Lat")

p_wind

#Short-wave radiation
summary_radiation <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(swr = mean(swr), .groups = 'drop')

p_radiation <- ggplot() + 
  geom_raster(data=summary_radiation, aes(x=lon, y=lat, fill=swr)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Radiation  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) +
  ggtitle("(g)") + map+
  xlab("Lon") + ylab("Lat")

p_radiation

#Surface Pressure
summary_sp <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(sp = mean(sp), .groups = 'drop')

p_sp <- ggplot() + 
  geom_raster(data=summary_sp, aes(x=lon, y=lat, fill=sp)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Pressure  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) +
  ggtitle("(h)") + map+
  xlab("Lon") + ylab("Lat")

p_sp

#Relative humidity
summary_rh <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(rh = mean(rh), .groups = 'drop')

p_rh <- ggplot() + 
  geom_raster(data=summary_rh, aes(x=lon, y=lat, fill=rh)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Humidity  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) + 
  ggtitle("(i)") + map+
  xlab("Lon") + ylab("Lat")

p_rh

#Precipitation
summary_precip <- data %>% 
  group_by(across(all_of(cols))) %>% 
  summarize(precip = mean(precip), .groups = 'drop')

p_precip <- ggplot() + 
  geom_raster(data=summary_precip, aes(x=lon, y=lat, fill=precip)) + 
  scale_fill_viridis_c(option = "viridis") +
  guides(fill = guide_legend(title = "Precip  ")) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.background = element_rect(fill = "white", color = "white", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2),
    legend.key.size = unit(0.4, 'cm'),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.title.align = 0,
    legend.position = "bottom", 
    legend.justification = "left",
    legend.direction = "horizontal"
  ) +
  ggtitle("") +
  ggtitle("(j)") + map+
  xlab("Lon") + ylab("Lat")

p_precip

# Create a layout matrix:
# The first row has all cells set to 1 (so p1 occupies the full top row),
# and the next rows assign the remaining plots in order.
layout_matrix <- rbind(
  c(1, 1, 1),       # Row 1: p1 spans all three columns
  c(2, 3, 4),       # Row 2: next three plots
  c(5, 6, 7),       # Row 3: following three plots
  c(8, 9, 10)       # Row 4: last three plots
)

# Arrange the plots using grid.arrange with the custom layout:
grid.arrange(p1, p_lightning, p_cape, p_cxp, p_temp, 
                  p_wind, p_radiation, p_sp, p_rh, p_precip,
                  layout_matrix = layout_matrix)

p <- grid.arrange(p1, p_lightning, p_cape, p_cxp, p_temp, 
                  p_wind, p_radiation, p_sp, p_rh, p_precip,
                  layout_matrix = layout_matrix)


ggsave(file.path(figure_root, "Figure_1.jpg"), plot = p, width = 12, height = 14, dpi = 750, limitsize = FALSE)
```



