---
title: "Figure 3"
author: "Charlotte Uden"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(forcats)
library(patchwork)
```

Fig 3. Density plots of model predictions and residuals. (a) Kernel density estimates of lightning strike rate, with color representing observed lightning strike rate (red) and lightning strike rates predicted from models N13 (yellow), G13 (green), and B13 (blue) . (b) Kernel density estimates of residuals (observed - predicted lightning strike rate) for models N13 (yellow), G13 (green), and B13 (blue). All models simulate lightning from six near-surface climate variables. 

Load configuration files
- location of the observed data (model drivers)
- location of table of model predictions exported from each model set's .Rmd file. 
- location to store the figure
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

config <- read_properties("/path/to/config/file/my_config.properties")  # path to your config file

# Extract values from config
data_root <- config[["data_root"]] 
table_root <- config[["table_root"]] 
figure_root <- config[["figure_root"]]

# Use the values to build full paths
Chen2021_file_path <- file.path(table_root, "Chen2021_models_predictions.csv")
lm_file_path <- file.path(table_root, "Linear_Model_predictions.csv")
glm_file_path <- file.path(table_root, "Gamma_GLM_predictions.csv")
bayes_file_path <- file.path(table_root, "bayes_gamma_model_predictions.csv")
monthly_drivers_file_path <- file.path(data_root, "era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")
```

```{r}
#read in the data
chen_predictions <- read.csv(Chen2021_file_path)[,2:6]
lm_predictions <- read.csv(lm_file_path)[,2:14]
glm_predictions <- read.csv(glm_file_path)[,2:14]
bayes_predictions <- read.csv(bayes_file_path)[,2:14]
```

Observed data - need to be split back into test and train sets because the above predictions are made from the test set. 
```{r}
data <- read.csv(monthly_drivers_file_path)
data <- data[2:13]
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```

Combine into one dataframe. Note that N13, G13, and B13 are used because they combine all size near-surface variables and are the best performing models within each of their respective model sets. 
```{r}
df <- data.frame(cbind(Observed = test$strikes,
                       N13 = lm_predictions$N13,
                       G13 = glm_predictions$G13,
                       B13 = bayes_predictions$y_pred_B13
                       ))

head(df)
```

Plot a:
```{r}
df_pivot <- df %>%
  pivot_longer(cols = c(Observed, N13, G13, B13), names_to = "source", values_to = "strike_rate")

df_pivot <- df_pivot %>%
  mutate(source = case_when(
    source == "N13" ~ "Linear Model (N13)",
    source == "G13" ~ "Gamma GLM (G13)",
    source == "B13" ~ "Gamma Bayesian (B13)",
    source == "Observed" ~ "Observed"))

df_pivot$source <- factor(df_pivot$source, levels = c(
  "Observed",
  "Linear Model (N13)",
  "Gamma GLM (G13)",
  "Gamma Bayesian (B13)"
))

p1 <- ggplot(df_pivot, aes(x = strike_rate, color = source, linetype = source, group=source)) +
  geom_density(alpha = 0.1, size = 0.8, bw=0.04) +
  labs(title = "(a)",
       x = expression("Lightning strike rate (strikes/km"^2*"/month)"),
       y = "Density") +
  scale_color_manual(values = c("Linear Model (N13)" = "purple4", 
                                "Gamma GLM (G13)" = "magenta3", 
                                "Gamma Bayesian (B13)" = "orange", 
                                "Observed" = "red")) +
  scale_linetype_manual(values = c("Linear Model (N13)" = "solid", 
                                   "Gamma GLM (G13)" = "dashed", 
                                   "Gamma Bayesian (B13)" = "longdash", 
                                   "Observed" = "dotted")) +
  guides(fill = "none") + 
  theme_minimal() +
  labs(fill = NULL, color = NULL, linetype=NULL) +
  theme(
     legend.position = c(0.99, 0.99),
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = "white", color = "darkgray", size = 0.2),
    plot.title = element_text(size = 13),
     axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)) 

p1

#kernal density estimate (KDE) estimates a probability density function 
#- the KDE method estiamtes some densites below (and above) the data. in this case, the smoothed curve extends below 0, even though the data do not contain negative values. 
# - KDE creates a continuous density curve which can extend beyond existing data points. 
# - KDE can estimate density values at points where no data exist- in this case, negative values. The PDFs estimated for the observed data and models G13 and B13 show negative values that do not exist. Model N13, however, does contain three negative predictions. 
```


Plot b:

```{r}

df <- df %>% 
  mutate(N13_residuals = Observed - N13) %>% 
  mutate(G13_residuals = Observed - G13) %>%
  mutate(B13_residuals = Observed - B13) 
  

df_pivot <- df %>%
  pivot_longer(cols = c(N13_residuals, G13_residuals, B13_residuals), names_to = "source", values_to = "strike_rate")

df_pivot <- df_pivot %>%
  mutate(source = case_when(
    source == "N13_residuals" ~ "Linear Model (N13)",
    source == "G13_residuals" ~ "Gamma GLM (G13)",
    source == "B13_residuals" ~ "Gamma Bayesian (B13)"))

df_pivot$source <- fct_inorder(df_pivot$source)  # Maintain original order

p2 <- ggplot(df_pivot, aes(x = strike_rate, color = source, linetype = source, group=source)) + 
  geom_density(alpha = 0.1, size = 0.8, bw=0.04) +
  labs(title = "(b)",
       x = expression("Residuals (strikes/km"^2*"/month)"),
       y = "Density") +
  scale_color_manual(values = c("Linear Model (N13)" = "purple4", 
                                "Gamma GLM (G13)" = "magenta3", 
                                "Gamma Bayesian (B13)" = "orange")) +
  scale_linetype_manual(values = c("Linear Model (N13)" = "solid", 
                                   "Gamma GLM (G13)" = "dashed", 
                                   "Gamma Bayesian (B13)" = "longdash")) +
  guides(fill = "none") +
  theme_minimal() +
  labs(fill = NULL, color = NULL, linetype=NULL) +
  theme(
    legend.position = c(0.99, 0.99),
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = "white", color = "darkgray", size = 0.2),
    plot.title = element_text(size = 13),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)) 

p1 | p2


ggsave(file.path(figure_root, "Figure_3.jpg"), plot = last_plot(), width = 12, height = 4, dpi = 750)
```


