---
title: "Figure 5"
author: "Charlotte Uden"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(gridExtra)
```
Fig. 4. Spatial comparison of observed and predicted lightning strike rates. Raster cells are colored by lightning strike rate (strikes/km2/month) averaged across six summers (2005 - 2010) of observed data and predictions from models N13, G13, and B13. See table 1 for model descriptions and definitions. All data are from the test set; white raster cells indicate latitude/longitude points not included in the test data due to the random splitting of the data into train (80%) and test (20%) sets.

Load configuration files
- location of the observed data (model drivers)
- location of table of model predictions exported from each model set's .Rmd file. 
- location to store the figure
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

config <- read_properties("/raid/cuden/config_files/lightning_config.properties")  # path to config file

# Extract values from config
data_root <- config[["data_root"]] 
table_root <- config[["table_root"]] 
figure_root <- config[["figure_root"]]

# Use the values to build full paths
Chen2021_file_path <- file.path(table_root, "Chen2021_models_predictions.csv")
lm_file_path <- file.path(table_root, "Linear_Model_predictions.csv")
glm_file_path <- file.path(table_root, "Gamma_GLM_predictions.csv")
bayes_file_path <- file.path(table_root, "bayes_gamma_model_predictions.csv")
monthly_drivers_file_path <- file.path(data_root, "era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")
```

```{r}
#read in the data
chen_predictions <- read.csv(Chen2021_file_path)[,2:6]
lm_predictions <- read.csv(lm_file_path)[,2:14]
glm_predictions <- read.csv(glm_file_path)[,2:14]
bayes_predictions <- read.csv(bayes_file_path)[,2:14]
```

Observed data - need to be split back into test and train sets because the above predictions are made from the test set. 
```{r}
data <- read.csv(monthly_drivers_file_path)
data <- data[2:13]
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```

Combine into one dataframe. Note that N13, G13, and B13 are used because they combine all size near-surface variables and are the best performing models within each of their respective model sets. 
```{r}
map_data <- data.frame(
  lat = test$lat,
  lon = test$lon,
  observed = test$strikes,
  N13 = lm_predictions$N13,
  G13 = glm_predictions$G13,
  B13 = bayes_predictions$y_pred_B13
)

head(map_data)
```

Plot observed and predicted strike rates over a map:
```{r}
# get new england and new york polygons from mapdata library :
states <- map_data("state")#turn state line map into data frame
northeast <- subset(states, region %in% c("vermont", "new hampshire", "connecticut", "maine", "rhode island", "massachusetts", "new york"))#subest northeastern states

map <- geom_polygon(data = northeast, aes(x=long, y = lat, group = group), fill = NA, color = "black", size=0.2) 

# Find the global min and max values across all datasets for consistent color scale
min_value <- min(c(map_data$observed, map_data$N13, map_data$G13, map_data$B13), na.rm = TRUE)
max_value <- max(c(map_data$observed, map_data$N13, map_data$G13, map_data$B13), na.rm = TRUE)

#get average observed value for each lat lon combo
avg_observed_data <- map_data %>%
  group_by(lat, lon) %>%
  summarize(avg_observed = mean(observed, na.rm = TRUE), .groups = "drop")

map_obs <- ggplot(avg_observed_data, 
                      aes(x = lon, y = lat, fill = avg_observed)) +
  geom_raster() +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    name = expression(Lightning~Strike~Rate),
    limits = c(min_value, max_value),  # Keep consistent scale
    guide = guide_colorbar(direction = "horizontal", title.position = "top")
  ) +
  xlab("Lon") +
  ylab("Lat") +
  theme_minimal() +
  ggtitle("Observed") +
   theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.position = c(0.05, 0.95),  # Position: top-left
    legend.justification = c("left", "top"),  # Align top-left corner
    legend.background = element_rect(fill = "white", color = "black", size = 0.1),
    plot.title = element_text(size = 13, hjust = 0),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  ) + map

# Map plots for each model
avg_pred_N13 <- map_data %>%
  group_by(lat, lon) %>%
  summarize(avg_pred = mean(N13, na.rm = TRUE), .groups = "drop")

map_pred_N13 <- ggplot(avg_pred_N13, 
                      aes(x = lon, y = lat, fill = avg_pred)) +
  geom_raster() +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    name = expression(Lightning~Strike~Rate),
    limits = c(min_value, max_value),  # Keep consistent scale
    guide = guide_colorbar(direction = "horizontal", title.position = "top")
  ) +
  xlab("Lon") +
  ylab("Lat") +
  theme_minimal() +
  ggtitle("Predicted: N13") +
   theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.position = c(0.05, 0.95),  # Position: top-left
    legend.justification = c("left", "top"),  # Align top-left corner
    legend.background = element_rect(fill = "white", color = "darkgray", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  ) + map

avg_pred_G13 <- map_data %>%
  group_by(lat, lon) %>%
  summarize(avg_observed = mean(G13, na.rm = TRUE), .groups = "drop")

map_pred_G13 <- ggplot(avg_pred_G13, 
                      aes(x = lon, y = lat, fill = avg_observed)) +
  geom_raster() +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    name = expression(Lightning~Strike~Rate),
    limits = c(min_value, max_value),  # Keep consistent scale
    guide = guide_colorbar(direction = "horizontal", title.position = "top")
  ) +
  xlab("Lon") +
  ylab("Lat") +
  theme_minimal() +
  ggtitle("Predicted: G13") +
   theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.position = c(0.05, 0.95),  # Position: top-left
    legend.justification = c("left", "top"),  # Align top-left corner
    legend.background = element_rect(fill = "white", color = "darkgray", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  ) + map

avg_pred_B13 <- map_data %>%
  group_by(lat, lon) %>%
  summarize(avg_observed = mean(B13, na.rm = TRUE), .groups = "drop")

map_pred_B13 <- ggplot(avg_pred_B13, 
                      aes(x = lon, y = lat, fill = avg_observed)) +
  geom_raster() +
  scale_fill_viridis_c(
    option = "plasma",  # Default Viridis color scale (use "cividis" or "plasma" if needed)
    direction = -1,
    name = expression(Lightning~Strike~Rate),
    limits = c(min_value, max_value),  # Keep consistent scale
    guide = guide_colorbar(direction = "horizontal", title.position = "top")
  ) +
  xlab("Lon") +
  ylab("Lat") +
  theme_minimal() +
  ggtitle("Predicted: B13") +
   theme(
     axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 13),
    legend.position = c(0.05, 0.95),  # Position: top-left
    legend.justification = c("left", "top"),  # Align top-left corner
    legend.background = element_rect(fill = "white", color = "darkgray", size = 0.2),
    plot.title = element_text(size = 13, hjust = 0),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    panel.grid.major = element_line(color = "darkgray", size = 0.2), 
    panel.grid.minor = element_line(color = "darkgray", size = 0.2)
  ) + map

(map_obs | map_pred_N13) / (map_pred_G13 | map_pred_B13)

ggsave(file.path(figure_root, "Figure_4.jpg"), plot = last_plot(), width = 12, height = 10, dpi = 750)
```



