---
title: "Linear Models"
author: "Charlotte Uden"
date: "2025-04-05"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(loo)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(gridExtra)
library(mapdata)
library(DescTools)
```

Load configuration files
- location of climate and lighting drivers
- location to store tables containing parameter estimates, model predictions, and performance metrics
```{r}
read_properties <- function(file_path) {
  lines <- readLines(file_path)
  lines <- lines[grepl("=", lines)]  # only keep lines with key-value pairs
  key_vals <- strsplit(lines, "=")
  props <- setNames(
    trimws(sapply(key_vals, `[`, 2)),
    trimws(sapply(key_vals, `[`, 1))
  )
  return(props)
}

config <- read_properties("/path/to/config/file/my_config.properties")  # path to your config file

# Extract values from config
data_root <- config[["data_root"]]
table_file_path <- config[["table_root"]]

# Use the values to build full paths
drivers_file_path <- file.path(data_root, "monthly_drivers.csv")
```

1. Read in the data

Make sure columns names are consistent with column names in this script: 
- lon (longitude)
- lat (latitude)
- year
- strikes (lightning strike rate, in flashes/km-2/month)
- cape (convective available potential energy, or CAPE in J kg-1)
- precip (precipitation in kg m-2 s-1)
- cxp (the product of CAPE and precipitation)
- tair (2 meter air temperature in degrees Celcius)
- wind (wind speed in m s-1)
- swr (shortwave radiation in W m-2)
- sp (surface pressure in Pa)
- rh (relative humidity expressed as a %)
Data should be summarized to monthly averages. 

```{r, echo=FALSE}
df <- read.csv(drivers_file_path)
#rename columns if necessary
#colnames(data) <- c("lon", "lat", "year", "strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")
```


2. Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]

```

#Standardize the data
```{r}
mu_cxp <- mean(train$cxp)
sigma_cxp <- sd(train$cxp)
mu_cape <- mean(train$cape)
sigma_cape <- sd(train$cape)
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)
#-----

train <- train %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh)

train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(train)

#use the min and max values from the train set to normalize the test set
test <- test %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) %>%
  mutate(lon = lon) %>%
  mutate(lat = lat)

test <- test[,c("lon", "lat", "strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

plot(test)

```

#Function to calculate S-score (Perkins et al. 2007)
```{r}
calculate_sscore <- function(obs, pred, nbins = 15, plot = FALSE) {
  # Determine common break points based on the range of both datasets
  min_val <- min(c(obs, pred))
  max_val <- max(c(obs, pred))
  breaks <- seq(min_val, max_val, length.out = nbins + 1)
  
  # Compute histograms for observed and predicted data using the same breaks
  hist_obs <- hist(obs, breaks = breaks, plot = FALSE)
  hist_pred <- hist(pred, breaks = breaks, plot = FALSE)
  
  # Optionally, plot the histograms for visual inspection
  if (plot) {
    plot(hist_obs, col = rgb(1, 0, 0, 0.5), main = "Histograms", xlab = "Value")
    plot(hist_pred, col = rgb(0, 0, 1, 0.5), add = TRUE)
    legend("topright", legend = c("Observed", "Predicted"),
           fill = c(rgb(1, 0, 0, 0.5), rgb(0, 0, 1, 0.5)))
  }
  
  # Convert counts to probabilities so that they sum to 1 - for each bin, take the frequency 
  #(counts in that bin) and divide by total counts to get a proportion - or probability. 
  prob_obs <- hist_obs$counts / sum(hist_obs$counts)
  prob_pred <- hist_pred$counts / sum(hist_pred$counts)
  
  # Calculate the overlap: the sum over bins of the minimum probability in each bin.
  sscore <- sum(pmin(prob_obs, prob_pred))
  
  return(sscore)
}
```


Generalized Linear models with family gaussian 
```{r}
#--------------------------
#CAPE
N1 <- lm(strikes ~ cape,
                 data = train)
a <- N1$coefficients[1]
b <- N1$coefficients[2]

y_pred_N1 <- a + b*test$cape

rmse_N1 <- Metrics::rmse(test$strikes, y_pred_N1)
cor_N1 <- cor(test$strikes, y_pred_N1)
sscore_N1 <- calculate_sscore(obs=test$strikes, pred=y_pred_N1, nbins = 15, plot = FALSE)

#--------------------------
#Cape x Precip
N2 <- lm(strikes ~ cxp,
                 data = train)
a <- N2$coefficients[1]
b <- N2$coefficients[2]

y_pred_N2 <- a + b*test$cxp

rmse_N2 <- Metrics::rmse(test$strikes, y_pred_N2)
cor_N2 <- cor(test$strikes, y_pred_N2)
sscore_N2 <- calculate_sscore(obs=test$strikes, pred=y_pred_N2, nbins = 15, plot = FALSE)

#--------------------------
#Relative Humidity
N3 <- lm(strikes ~ rh,
                 data = train)
a <- N3$coefficients[1]
b <- N3$coefficients[2]

y_pred_N3 <- a + b*test$rh

rmse_N3 <- Metrics::rmse(test$strikes, y_pred_N3)
cor_N3 <- cor(test$strikes, y_pred_N3)
sscore_N3 <- calculate_sscore(obs=test$strikes, pred=y_pred_N3, nbins = 15, plot = FALSE)

#--------------------------
#Short Wave Radition
N4 <- lm(strikes ~ swr,
                 data = train)
a <- N4$coefficients[1]
b <- N4$coefficients[2]

y_pred_N4 <- a + b*test$swr

rmse_N4 <- Metrics::rmse(test$strikes, y_pred_N4)
cor_N4 <- cor(test$strikes, y_pred_N4)
sscore_N4 <- calculate_sscore(obs=test$strikes, pred=y_pred_N4, nbins = 15, plot = FALSE)

#--------------------------
#Temperature
N5 <- lm(strikes ~ tair,
                 data = train)
a <- N5$coefficients[1]
b <- N5$coefficients[2]

y_pred_N5 <- a + b*test$tair

rmse_N5 <- Metrics::rmse(test$strikes, y_pred_N5)
cor_N5 <- cor(test$strikes, y_pred_N5)
sscore_N5 <- calculate_sscore(obs=test$strikes, pred=y_pred_N5, nbins = 15, plot = FALSE)

#--------------------------
#Surface Pressure
N6 <- lm(strikes ~ sp,
                 data = train)
a <- N6$coefficients[1]
b <- N6$coefficients[2]

y_pred_N6 <- a + b*test$sp

rmse_N6 <- Metrics::rmse(test$strikes, y_pred_N6)
cor_N6 <- cor(test$strikes, y_pred_N6)
sscore_N6 <- calculate_sscore(obs=test$strikes, pred=y_pred_N6, nbins = 15, plot = FALSE)

#--------------------------
#Precipitation
N7 <- lm(strikes ~ precip,
                 data = train)
a <- N7$coefficients[1]
b <- N7$coefficients[2]

y_pred_N7 <- a + b*test$precip

rmse_N7 <- Metrics::rmse(test$strikes, y_pred_N7)
cor_N7 <- cor(test$strikes, y_pred_N7)
sscore_N7 <- calculate_sscore(obs=test$strikes, pred=y_pred_N7, nbins = 15, plot = FALSE)

#--------------------------
#Precipitation
N8 <- lm(strikes ~ wind,
                 data = train)
a <- N8$coefficients[1]
b <- N8$coefficients[2]

y_pred_N8 <- a + b*test$wind

rmse_N8 <- Metrics::rmse(test$strikes, y_pred_N8)
cor_N8 <- cor(test$strikes, y_pred_N8)
sscore_N8 <- calculate_sscore(obs=test$strikes, pred=y_pred_N8, nbins = 15, plot = FALSE)

#--------------------------
#swr + tair
N9 <- lm(strikes ~ swr + tair,
                 data = train)
a <- N9$coefficients[1]
b <- N9$coefficients[2]
c <- N9$coefficients[3]

y_pred_N9 <- a + b*test$swr + c*test$tair

rmse_N9 <- Metrics::rmse(test$strikes, y_pred_N9)
cor_N9 <- cor(test$strikes, y_pred_N9)
sscore_N9 <- calculate_sscore(obs=test$strikes, pred=y_pred_N9, nbins = 15, plot = FALSE)

#--------------------------
#swr + tair + rh
N10 <- lm(strikes ~ swr + tair + rh,
                 data = train)
a <- N10$coefficients[1]
b <- N10$coefficients[2]
c <- N10$coefficients[3]
d <- N10$coefficients[4]

y_pred_N10 <- a + b*test$swr + c*test$tair + d*test$rh

rmse_N10 <- Metrics::rmse(test$strikes, y_pred_N10)
cor_N10 <- cor(test$strikes, y_pred_N10)
sscore_N10 <- calculate_sscore(obs=test$strikes, pred=y_pred_N10, nbins = 15, plot = FALSE)

#--------------------------
#swr + tair + rh + wind
N11 <- lm(strikes ~ swr + tair + rh + wind,
                 data = train)
a <- N11$coefficients[1]
b <- N11$coefficients[2]
c <- N11$coefficients[3]
d <- N11$coefficients[4]
e <- N11$coefficients[5]

y_pred_N11 <- a + b*test$swr + c*test$tair + d*test$rh +e*test$wind

rmse_N11 <- Metrics::rmse(test$strikes, y_pred_N11)
cor_N11 <- cor(test$strikes, y_pred_N11)
sscore_N11 <- calculate_sscore(obs=test$strikes, pred=y_pred_N11, nbins = 15, plot = FALSE)

#--------------------------
#swr + tair + rh + wind + precip
N12 <- lm(strikes ~ swr + tair + rh + wind + precip,
                 data = train)
a <- N12$coefficients[1]
b <- N12$coefficients[2]
c <- N12$coefficients[3]
d <- N12$coefficients[4]
e <- N12$coefficients[5]
f <- N12$coefficients[6]

y_pred_N12 <- a + b*test$swr + c*test$tair + d*test$rh +e*test$wind + f*test$precip

rmse_N12 <- Metrics::rmse(test$strikes, y_pred_N12)
cor_N12 <- cor(test$strikes, y_pred_N12)
sscore_N12 <- calculate_sscore(obs=test$strikes, pred=y_pred_N12, nbins = 15, plot = FALSE)

#--------------------------
#swr + tair + rh + wind + precip +sp
N13 <- lm(strikes ~ swr + tair + rh + wind + precip +sp,
                 data = train)
a <- N13$coefficients[1]
b <- N13$coefficients[2]
c <- N13$coefficients[3]
d <- N13$coefficients[4]
e <- N13$coefficients[5]
f <- N13$coefficients[6]
g <- N13$coefficients[7]

y_pred_N13 <- a + b*test$swr + c*test$tair + d*test$rh +e*test$wind + f*test$precip + g*test$sp

rmse_N13 <- Metrics::rmse(test$strikes, y_pred_N13)
cor_N1 <- cor(test$strikes, y_pred_N13)
sscore_N13 <- calculate_sscore(obs=test$strikes, pred=y_pred_N13, nbins = 15, plot = FALSE)
```

#Export table of predictions
```{r}
predictions <- data.frame(cbind(N1 = y_pred_N1, N2 = y_pred_N2, N3 = y_pred_N3, N4 = y_pred_N4, N5 = y_pred_N5, N6 = y_pred_N6, N7 = y_pred_N7, N8 = y_pred_N8, N9 = y_pred_N9, N10 = y_pred_N10, N11 = y_pred_N11, N12 = y_pred_N12, N13 = y_pred_N13))

print(predictions)

write.csv(predictions, file.path(table_file_path, "Linear_Model_predictions.csv"))
```

#Export table of performance metrics
```{r}
model_name <- c("N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9", "N10", "N11", "N12", "N13")
fitting_function <- c("glm", "glm", "glm", "glm", "glm",  "glm", "glm", "glm", "glm", "glm",  "glm", "glm", "glm")
distribution <- c("gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian", "gaussian")
group_name <- rep("Linear Model", 13)
predictors <- c("CAPE", "CAPE x P", "RH", "SWR", "T", "SP", "P", "W", "SWR + T", "SWR + T + RH", "SWR + T + RH + W", "SWR + T + RH + W + + P", "SWR + T + RH + W + P + SP")

rmse <- c(rmse_N1, rmse_N2, rmse_N3, rmse_N4, rmse_N5, rmse_N6, rmse_N7, rmse_N8, rmse_N9, rmse_N10, rmse_N11, rmse_N12, rmse_N13)

cor <- c(cor(test$strikes, y_pred_N1), cor(test$strikes, y_pred_N2), cor(test$strikes, y_pred_N3), cor(test$strikes, y_pred_N4), cor(test$strikes, y_pred_N5), cor(test$strikes, y_pred_N6), cor(test$strikes, y_pred_N7), cor(test$strikes, y_pred_N8), cor(test$strikes, y_pred_N9), cor(test$strikes, y_pred_N10), cor(test$strikes, y_pred_N11), cor(test$strikes, y_pred_N12), cor(test$strikes, y_pred_N13))

sscore <- c(sscore_N1, sscore_N2, sscore_N3, sscore_N4, sscore_N5, sscore_N6, sscore_N7, sscore_N8, sscore_N9, sscore_N10, sscore_N11, sscore_N12, sscore_N13)

performance <- data.frame(cbind(model_name, fitting_function, distribution, predictors, group_name, rmse, cor, sscore))

print(performance)

write.csv(performance, file.path(table_file_path, "Linear_Model_performance.csv"))

```

#Loop through models to get parameter estimates and calculate 95% condience interval (Estimate ± 1.96 × SE)

```{r}
# Store all models and their labels
model_list <- list(N1, N2, N3, N4, N5, N6, N7, N8, N9, N10, N11, N12, N13)
model_labels <- paste("N", 1:13)
predictor_list <- c("CAPE", "CAPE x P", "RH", "SWR", "T", "SP", "P", "W",
                    "SWR + T", "SWR + T + RH", "SWR + T + RH + W",
                    "SWR + T + RH + W + P", "SWR + T + RH + W + P + SP")
functional_forms <- c("a + b * CAPE","a + b * CAPE * P","a + b * RH","a + b * SWR","a + b * T","a + b * SP","a + b * P","a + b * W","a + b * SWR + c * T","a + b * SWR + c * T + d * RH","a + b * SWR + c * T + d * RH + e * W","a + b * SWR + c * T + d * RH + e * W + f * P","a + b * SWR + c * T + d * RH + e * W + f * P + g * SP")



# Initialize empty data frame
param_table <- data.frame(model_label = character(),
                          predictors = character(),
                          a = character(), b = character(), c = character(),
                          d = character(), e = character(), f = character(), g = character(),
                          stringsAsFactors = FALSE)

# Loop over each model to extract coefficients and standard errors
for (i in seq_along(model_list)) {
  model_summary <- summary(model_list[[i]])$coefficients
  coefs <- model_summary[, "Estimate"]
  ses <- model_summary[, "Std. Error"]
  
  # Pad to length 7
  coefs_padded <- c(coefs, rep(NA, 7 - length(coefs)))
  ses_padded <- c(ses, rep(NA, 7 - length(ses)))
  
  # Format estimate ± 1.96*SE
  format_ci <- function(est, se) {
    if (is.na(est) || is.na(se)) return(NA)
    error <- 1.96 * se
    sprintf("%.3f ± %.3f", est, error)
  }
  
  param_table <- rbind(param_table, data.frame(
    model_label = model_labels[i],
    predictors = predictor_list[i],
    functional_form = functional_forms[i],
    a = format_ci(coefs_padded[1], ses_padded[1]),
    b = format_ci(coefs_padded[2], ses_padded[2]),
    c = format_ci(coefs_padded[3], ses_padded[3]),
    d = format_ci(coefs_padded[4], ses_padded[4]),
    e = format_ci(coefs_padded[5], ses_padded[5]),
    f = format_ci(coefs_padded[6], ses_padded[6]),
    g = format_ci(coefs_padded[7], ses_padded[7])
  ))
}

print(param_table)

# Export the final table
write.csv(param_table, file.path(table_file_path, "Table_3.csv"), row.names = FALSE)

```


